Long term cloud enhancement events of global solar irradiance over Thessaloniki, Greece

### 2024-06-12 article/article.Rmd
Commit 2024-06-12 19:11

We studied the enhancement of the global horizontal irradiance (GHI), over Thessaloniki, 
for the period 1993 -- 2023, on one minute average records.
We identify the cloud enhancement events (CE), by creating an appropriate clear sky
irradiance reference with the use of a radiation transfer model, and aerosol optical 
depth (AOD) data, from the AERONET, and a collocated Brewer photospectrometer.
We found that there is a trend of CE cases of
$`r round(yearlyCE[var == "GLB_diff.N", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$,
and a
mean total energy change, of the CE events by
$`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kj}/\text{year}$
with the peak of the CE events are observed during May and June.
An analysis of the total duration of CE events, showed that durations longer than 
5 minutes are very rare, with exceptions lasting over an hour.
We have detected enhancements over the total solar irradiance, without exceeding the
maximum values reported for other sites, with more favorable CE conditions.

@@

For the detection of CE cases we established a baseline of irradiance above which we
characterized each data point as an enhancement event and calculated the over
irradiance (OIR). The OIR is defined as the irradiance difference of the measured
one-minute $\text{GHI}_i$ from the $\text{GHI}_\text{ref}$ corresponding to
cloud-free atmosphere. 
<!-- CE identification criterion ($\text{OIR}_i =
\text{GHI}_i - \text{GHI}_\text{CSlim}$), as defined in
Equation\nobreakspace\ref{eq:CE4}. -->
First, we tested two simple approaches for the determination of
$\text{GHI}_\text{ref}$: The Haurwitz's model [@Haurwitz1945], which is a simple
clear sky model and was already adjusted and applied to our data [@Natsis2023], and
the total solar irradiance (TSI) at the top of the atmosphere, adjusted for the
Sun-Earth distance.  We have tested both methods by using an appropriate relative
threshold and/or an additional constant offset.  The initial results showed that we
can detect a big portion of the CE events. However, by inspecting the daily plots of
irradiance it became evident that changes in atmospheric conditions introduced
numerous false positive and false negative results.  The main reason for these
discrepancies is the variability of the effects of aerosols and water vapor which
were not taken into account in the two simple methods.  To produce a more
representative reference  we included the effects of these factors through variables
using a radiative transfer model (RTM).  The applied methodology is discussed in
section\nobreakspace{}\ref{rtmcs}

@@

<!-- Aerosol change -->
The above clear sky reference values are based on the climatological AOD and WC,
hence they cannot describe accurately the long term variation of
$\text{GHI}_\text{ref}$ due to changes in the two atmospheric constituents, mainly
AOD.  As reported by @Natsis2023, there is a long term brightening effect in the GHI
data of Thessaloniki for the period 1993 -- 2023, which for clear-sky data was
attributed to long-term changes in aerosol effects.  Therefore, an adjustment of the
$\text{GHI}_\text{ref}$ used in this study is necessary. As AERONET data start only
in 2003, we used for the period 1993 -– 2005 estimates of changes in AOD at
$340\,\text{nm}$ derived from a collocated Brewer spectrophotometer derived from two
sources. For 1993 -- 2005 [@Kazadzis2007] to calculate the trend in
$\text{GHI}_\text{ref}$ during this period.

@@

Following the application of the above discussed methodology to the entire dataset 
(`r round(ST_total$wattGLB.N / 10^6,1)` million of one-minute GHI measurements),
$`r round(100*ST_total$GLB_diff.N_pos/ST_total$wattGLB.N, 3)`\,\%$ were identified as CE events and
$`r round(100*ST_extreme_total$GLB_diff.N_pos/ST_total$wattGLB.N, 3)`\,\%$ as ECE events.
The highest recorded GHI due to CE was
$`r round(ST_daily[which.max(wattGLB.max), wattGLB.max],1)`\,\text{W}/\text{m}^2$
on `r strftime(ST_daily[which.max(wattGLB.max), Date], "%d\\ %B %Y")`, corresponding
to OIR of $`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 1)`\,\text{W}/\text{m}^2$.
<!-- The absolute strongest CE event had an OIR of 
<!-- $`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 2)`\,\text{W}/\text{m}^2$ on
<!-- `r strftime(ST_daily[which.max(GLB_diff.max), Date], "%d\\ %B %Y")`. -->
The stronger CE event of
$`r round(100*ST_daily[which.max(GLB_ench.max), GLB_ench.max],1)`\,\%$ above the 
clear sky threshold was observed on 
`r strftime(ST_daily[which.max(GLB_ench.max), Date], "%d\\ %B %Y")`.
In the following sections we are discussing the long-term trends and variability of the CE events as well as of the corresponding OIR and excess irradiation.

@@

Another interesting aspect of the CE events is their distribution within the year.
Figure\nobreakspace{}\ref{fig:relative-month-occurrences} shows the monthly box and
whisker plot of the CE number of occurrence normalized with the highest median value,
that occurs in June, depicting a clear seasonal cycle.  Although CE events are
present  throughout the year, the most active period  is during May and June.  During
the winter (December -- February), the number of CE cases is about $25\,\%$ of the
maximum, while in the intermediate months, the number of occurrences gradually ramps
between the maximum and minimum.  This seasonality is a combined effect of different
factors, among them the types of clouds, their frequency of occurrence the seasonally
varying relative position of the sun, as well as the local landscape characteristics
that may influence the formation of clouds.  Unfortunately, lack of detailed data on
cloud formation, type and location is not allowing further analysis. The interannual
variability of the monthly CE events is quite high as manifested by the large monthly
extremes, especially in the summer.

```{r relative-month-occurrences, echo=FALSE, fig.cap="Seasonal variability of the number of CE events normalized to the maximum occurring in June, in the form of a box and whisker plot. The monthly values have been also normalized to the relative abundance of valid GHI observations. The box contains the data between the lower $25\\,\\%$ and the upper $75\\,\\%$ percentiles,The thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values and the solid circles are outliers."}
knitr::include_graphics("../images/clim-CE-month-norm-MAX-median-N-MW-1.png")
```
<!--
```{r echo=FALSE, fig.cap="Distribution of the number of CE cases"}
knitr::include_graphics("../images/climCEmonth-1.png")
```
--->

@@

The distribution of the number of CE events as a function of OIR is shown in
Figure\nobreakspace{}\ref{fig:ovir-distribution}. Apparently, there is an inverse
relation between the frequency of CE events  and OIR with an exponential-like
decline. This is expected, as the stronger the CE events become, the rarer are the
conditions favoring the occurrence of CE events. For the majority (over $62\,\%$) of
the CE events the OIR is below the long term average of
$`r round(mean(ST_E_yearly$GLB_diff.mean), 1)`\,\text{W}/\text{m}^2$,
while about $8.1\,\%$ of the events correspond to OIR larger than
$100\,\text{W}/\text{m}^2$ and up to highest value of $412.4\,\text{W}/\text{m}^2$.
This distribution is indicative of the magnitude and the probability of the expected
CE events over Thessaloniki.  Similar distribution of CE events, albeit with larger
OIR values, has been reported by @Vamvakas2020, for the city of Patras.This site is
located $2.5^\circ$ south of Thessaloniki and is exposed to air masses coming mainly
from the eastern Mediterranean resulting in different cloud patterns, that may affect
the characteristics and magnitude of the CE events.

@@

The relation between the duration and the mean OIR intensity of the groups have also been
studied (Figure\nobreakspace{}\ref{fig:group-2d}). We observed that the GCE events
tend to have either long duration, or large intensity. Events with strong enhancement
and long duration are very rare. Similar results on this relation, have been
reported by @Zhang2018, on a study using a far higher sampling rate than ours.

@@

<!-- ```{r echo=FALSE, fig.cap="Relation of mean over irradiance and CE group duration"} -->
<!-- knitr::include_graphics("../images/groups-4.png") -->
<!-- ``` -->

<!-- ```{r echo=FALSE, fig.cap="Relation of maximun over irradiance and CE group duration"} -->
<!-- knitr::include_graphics("../images/groups-8.png") -->
<!-- ``` -->

@@

An aspect of the CE events that is commonly reported and has some significance
on the solar energy production infrastructure are the extreme CE events (ECE). Where
solar irradiance exceeds the expected irradiance on top of the atmosphere
(Equation\nobreakspace{}\ref{eq:ECE}).
Analogous to Figure\nobreakspace{}\ref{fig:relative-month-occurancies} we have
computed the distribution of the number of occurrences of ECE events by month in
Figure\nobreakspace{}\ref{fig:relative-month-occurancies-ECE}. The most active period
for ECE events is in the spring and the start of the summer (March -- June), followed
by a period in the late fall (September and October).  This is probably related to 
the weather characteristics of these periods, where there are continuous alternations
between clear sky periods and clouds.

@@

By creating a clear sky approximation of the GHI, which represents the long- and
short-term variation of the expected clear sky GHI, we were able to identify cases of 
CE events.
After analyzing the CE cases, we found an increase of 
$`r round(yearlyCE[var == "GLB_diff.N", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$,
with the
mean annual total energy of the CE events increasing with a rate of
$`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kj}/\text{year}$.
The most active season of CE events over Thessaloniki is concentrated on early
summer, on May and June.

>>>>

We studied the enhancement of the global horizontal irradiance (GHI), over Thessaloniki, 
for the period 1993 -- 2023, on one minute average records.
We identify the cloud enhancement events (CE), by creating an appropriate clear sky
irradiance reference with the use of a radiation transfer model, and aerosol optical 
depth (AOD) data, from the AERONET, and a collocated Brewer photospectrometer.
We found that there is a trend of CE cases of
$`r round(yearlyCE[var == "GLB_diff.N", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$,
and a
mean total energy change, of the CE events by
$`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$
with the peak of the CE events are observed during May and June.
An analysis of the total duration of CE events, showed that durations longer than 
5 minutes are very rare, with exceptions lasting over an hour.
We have detected enhancements over the total solar irradiance, without exceeding the
maximum values reported for other sites, with more favorable CE conditions.

@@

For the detection of CE cases we established a baseline of irradiance above which we
characterized each data point as an enhancement event and calculated the over
irradiance (OIR). The OIR is defined as the irradiance difference of the measured
one-minute $\text{GHI}_i$ from the $\text{GHI}_\text{ref}$ corresponding to
cloud-free atmosphere. 
<!-- CE identification criterion ($\text{OIR}_i =
\text{GHI}_i - \text{GHI}_\text{CSlim}$), as defined in
Equation\nobreakspace\ref{eq:CE4}. -->
First, we tested two simple approaches for the determination of
$\text{GHI}_\text{ref}$: The Haurwitz's model [@Haurwitz1945], which is a simple
clear sky model and was already adjusted and applied to our data [@Natsis2023], and
the total solar irradiance (TSI) at the top of the atmosphere, adjusted for the
Sun-Earth distance.  We have tested both methods by using an appropriate relative
threshold and/or an additional constant offset.  The initial results showed that we
can detect a big portion of the CE events. However, by inspecting the daily plots of
irradiance it became evident that changes in atmospheric conditions introduced
numerous false positive and false negative results.  The main reason for these
discrepancies is the variability of the effects of aerosols and water vapor which
were not taken into account in the two simple methods.  To produce a more
representative reference  we included the effects of these factors through variables
using a radiative transfer model (RTM).  The applied methodology is discussed in
section\nobreakspace{}\ref{rtmcs}..

@@

<!-- Aerosol change -->
The above clear sky reference values are based on the climatological AOD and WC,
hence they cannot describe accurately the long term variation of
$\text{GHI}_\text{ref}$ due to changes in the two atmospheric constituents, mainly
AOD.  As reported by @Natsis2023, there is a long term brightening effect in the GHI
data of Thessaloniki for the period 1993 -- 2023, which for clear-sky data was
attributed to long-term changes in aerosol effects.  Therefore, an adjustment of the
$\text{GHI}_\text{ref}$ used in this study is necessary. As AERONET data start only
in 2003, we used for the period 1993 -- 2005 estimates of changes in AOD at
$340\,\text{nm}$ derived from a collocated Brewer spectrophotometer [@Kazadzis2007]
to calculate the trend in $\text{GHI}_\text{ref}$ during this period.

@@

Following the application of the above discussed methodology to the entire dataset 
(`r round(ST_total$wattGLB.N / 10^6,1)` million of one-minute GHI measurements),
$`r round(100*ST_total$GLB_diff.N_pos/ST_total$wattGLB.N, 2)`\,\%$ were identified as CE events and
$`r round(100*ST_extreme_total$GLB_diff.N_pos/ST_total$wattGLB.N, 3)`\,\%$ as ECE events.
The highest recorded GHI due to CE was
$`r round(ST_daily[which.max(wattGLB.max), wattGLB.max],1)`\,\text{W}/\text{m}^2$
on `r strftime(ST_daily[which.max(wattGLB.max), Date], "%d\\ %B %Y")`, corresponding
to OIR of $`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 1)`\,\text{W}/\text{m}^2$.
<!-- The absolute strongest CE event had an OIR of 
<!-- $`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 2)`\,\text{W}/\text{m}^2$ on
<!-- `r strftime(ST_daily[which.max(GLB_diff.max), Date], "%d\\ %B %Y")`. -->
The stronger CE event of
$`r round(100*ST_daily[which.max(GLB_ench.max), GLB_ench.max],1)`\,\%$ above the 
clear sky threshold was observed on 
`r strftime(ST_daily[which.max(GLB_ench.max), Date], "%d\\ %B %Y")`.
In the following sections we are discussing the long-term trends and variability of the CE events as well as of the corresponding OIR and excess irradiation.

@@

Another interesting aspect of the CE events is their distribution within the year.
Figure\nobreakspace{}\ref{fig:relative-month-occurrences} shows the monthly box and
whisker plot of the CE number of occurrence normalized with the highest median value
that occurs in June, depicting a clear seasonal cycle.  Although CE events are
present  throughout the year, the most active period  is during May and June.  During
winter (December -- February), the number of CE cases is about $25\,\%$ of the
maximum, while in the intermediate months the number of occurrences gradually ramps
between the maximum and minimum.  This seasonality is a combined effect of different
factors, among them the types of clouds, their frequency of occurrence the seasonally
varying relative position of the sun, as well as the local landscape characteristics
that may influence the formation of clouds.  Unfortunately, lack of detailed data on
cloud formation, type and location is not allowing further analysis. The interannual
variability of the monthly CE events is quite high as manifested by the large monthly
extremes, especially in the summer.

```{r relative-month-occurrences, echo=FALSE, fig.cap="Seasonal variability of the number of CE events normalized to the maximum occurring in June, in the form of a box and whisker plot. The monthly values have been also normalized to the relative abundance of valid GHI observations. The box contains the data between the lower $25\\,\\%$ and the upper $75\\,\\%$ percentiles. The thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values and the solid circles are outliers."}
knitr::include_graphics("../images/clim-CE-month-norm-MAX-median-N-MW-1.png")
```
<!--
```{r echo=FALSE, fig.cap="Distribution of the number of CE cases"}
knitr::include_graphics("../images/climCEmonth-1.png")
```
--->

@@

The distribution of the number of CE events as a function of OIR is shown in
Figure\nobreakspace{}\ref{fig:ovir-distribution}. Apparently, there is an inverse
relation between the frequency of CE events  and OIR with an exponential-like
decline. This is expected, as the stronger the CE events become, the rarer are the
conditions favoring the occurrence of CE events. For the majority (over $62\,\%$) of
the CE events the OIR is below the long term average of
$`r round(mean(ST_E_yearly$GLB_diff.mean), 1)`\,\text{W}/\text{m}^2$,
while about $8.1\,\%$ of the events correspond to OIR larger than
$100\,\text{W}/\text{m}^2$ and up to highest value of $412.4\,\text{W}/\text{m}^2$.
This distribution is indicative of the magnitude and the probability of the expected
CE events over Thessaloniki.  Similar distribution of CE events, albeit with larger
OIR values, has been reported by @Vamvakas2020 for the city of Patras. This site is
located $2.5^\circ$ south of Thessaloniki and is exposed to air masses coming mainly
from the eastern Mediterranean resulting in different cloud patterns, that may affect
the characteristics and magnitude of the CE events.

@@

The relation between the duration of the CE groups and the excess irradiation of the
groups has also been investigated (Figure\nobreakspace{}\ref{fig:group-2d}).
Evidently, evens of high excess irradiation have small duration and vice versa. The
vast majority of grouped events are associated with small excess irradiation (e.g.
$<5\,\text{kJ}/\text{m}^2$) and small duration (e.g. $<5\,\text{min}$) while groups
with strong excess irradiation and long duration are rare.  Similar results on this
relation, have been reported by @Zhang2018, on a study using a far higher sampling
rate than ours.

@@

<!-- ```{r echo=FALSE, fig.cap="Relation of mean over irradiance and CE group duration"} -->
<!-- knitr::include_graphics("../images/groups-4.png") -->
<!-- ``` -->
<!-- ```{r echo=FALSE, fig.cap="Relation of maximun over irradiance and CE group duration"} -->
<!-- knitr::include_graphics("../images/groups-8.png") -->
<!-- ``` -->

@@

An aspect of the CE events that is commonly reported and has some significance
on the solar energy production infrastructure are the extreme CE events (ECE). Where
solar irradiance exceeds the expected irradiance on top of the atmosphere
(Equation\nobreakspace{}\ref{eq:ECE}).
Analogous to Figure\nobreakspace{}\ref{fig:relative-month-occurrences} we have
computed the distribution of the number of occurrences of ECE events by month in
Figure\nobreakspace{}\ref{fig:relative-month-occurancies-ECE}. The most active period
for ECE events is in the spring and the start of the summer (March -- June), followed
by a period in the late fall (September and October).  This is probably related to 
the weather characteristics of these periods, where there are continuous alternations
between clear sky periods and clouds.

@@

By creating a clear sky approximation of the GHI, which represents the long- and
short-term variation of the expected clear sky GHI, we were able to identify cases of 
CE events.
After analyzing the CE cases, we found an increase of 
$`r round(yearlyCE[var == "GLB_diff.N", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$,
with the
mean annual total energy of the CE events increasing with a rate of
$`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)`\pm
`r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$.
The most active season of CE events over Thessaloniki is concentrated on early
summer, on May and June.
